---
title: "MD_Analysis_race"
author: "Leng Seong Che & Casey F. Breen"
date: "2026-01-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(here)
library(readr)
library(dplyr)
library(ggplot2)
library(MASS) 
library(lubridate)
library(caret)
library(splines)
library(glmnet)
library(tidyr)
library(kableExtra)
library(brms)
```

```{r}
MD_aggregate_month_cause <- read_csv(here("data", "Mortality data", "Aggregate cause-specific files", "MD_aggregate_month-cause - v1.csv"))
MD_pop <- read_csv(here("data", "Denominator data", "Census counts", "MD_population - v1.csv"))
MD_soldiers <- read_csv(here("data", "Denominator data", "Drafted soldier counts", "MD_soldiers - v1.csv"))
```

```{r}
death_month <- MD_aggregate_month_cause %>%
  mutate(year = as.integer(year),
         month = as.integer(month),
         ym = as.Date(sprintf("%04d-%02d-01", year, month))) %>%
  group_by(ym,race) %>%
  summarise(monthly_deaths = sum(deaths, na.rm = TRUE), .groups = "drop")

pandemic_start <- as.Date("1918-03-01")
pandemic_end   <- as.Date("1920-05-01")

ggplot(death_month, aes(x = ym, y = monthly_deaths, color = race)) +
  geom_rect(
    data = data.frame(xmin = pandemic_start, xmax = pandemic_end, ymin = -Inf, ymax = Inf),
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = "orange", alpha = 0.12, inherit.aes = FALSE
  ) +
  geom_line(linewidth = 0.8) +
  #geom_smooth(method = "loess", span = 0.2, se = FALSE) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    x = "Month",
    y = "Number of deaths",
    title = "All-Cause Death Counts per Month in Rural Maryland, by Race",
    color = "Race"
  ) +
  theme_minimal()
```

# Weather 

## Humidity

```{r}
MD_weather <- read_csv(here("data", "Weather data", "MD_weather - v1.csv"))
```

```{r}
precip_month <- MD_weather %>%
  mutate(ym = as.Date(sprintf("%04d-%02d-01", year, month))) %>%
  group_by(ym) %>%
  summarise(precip = mean(precipitation1, na.rm = TRUE), .groups = "drop")

precip_month <- death_month %>% 
  mutate(year = year(ym), month = month(ym))%>% 
  left_join(precip_month, by = "ym")

precip_month <- precip_month %>%                    
  group_by(month) %>%
  mutate(
    mean_precip_month = mean(precip, na.rm = TRUE),
    sd_precip_month   = sd(precip,   na.rm = TRUE),
    high_precip = ifelse(precip > mean_precip_month, 1, 0),
    very_high_precip = ifelse(precip > mean_precip_month + sd_precip_month, 1, 0)
  ) %>%
  ungroup()
```

## Temperature

```{r}
temp_month <- MD_weather %>%
  mutate(ym = as.Date(sprintf("%04d-%02d-01", year, month))) %>%
  group_by(ym) %>%
  summarise(monthly_mean_temp = mean(temperature1, na.rm = TRUE), 
            monthly_mean_avg_temp = mean(temperature2, na.rm = TRUE),
            .groups = "drop")

temp_month <- death_month %>% 
  mutate(year = year(ym), month = month(ym))%>% 
  left_join(temp_month, by = "ym")

temp_month <- temp_month %>%                    
  group_by(month) %>%
  mutate(
    mean_temp_month = mean(monthly_mean_temp, na.rm = TRUE),
    mean_avg_temp_month = mean(monthly_mean_avg_temp, na.rm = TRUE),
    sd_temp_month   = sd(monthly_mean_temp,   na.rm = TRUE),
    sd_avg_temp_month = sd(monthly_mean_avg_temp, na.rm = TRUE),
    high_temp = ifelse(monthly_mean_temp > mean_temp_month, 1, 0),
    very_high_temp = ifelse(monthly_mean_temp > mean_temp_month + sd_temp_month, 1, 0),
    high_avg_temp = ifelse(monthly_mean_avg_temp > mean_avg_temp_month, 1, 0),
    very_high_avg_temp = ifelse(monthly_mean_avg_temp > mean_avg_temp_month + sd_avg_temp_month, 1, 0)
  ) %>%
  ungroup()
```

# CDR

```{r}
MD_pop_year <- MD_pop %>%
  group_by(year, race) %>%
  summarise(pop = sum(population, na.rm=TRUE), .groups = "drop")

decimal_year <- function(d) { d <- as.Date(d); year(d) + (month(d) - 0.5)/12 }

all_months <- sort(unique(as.Date(death_month$ym)))
all_dec    <- decimal_year(all_months)

pop_monthly_race <- MD_pop_year %>%
  arrange(race, year) %>%
  group_by(race) %>%
  do({
    census_years <- .$year
    census_pops  <- .$pop
    data.frame(
      ym = all_months,
      race = unique(.$race),
      pop_interp = exp(approx(
        x = census_years,
        y = log(census_pops),
        xout = all_dec,
        rule = 2
      )$y)
    )
  }) %>%
  ungroup()


death_month <- death_month %>%
  left_join(pop_monthly_race, by = c("ym", "race")) %>%
  mutate(crude_rate_100k = monthly_deaths / pop_interp * 100000)

ggplot(death_month, aes(x = ym, y = crude_rate_100k, color = race)) +
  geom_line() +
  labs(
    x = "Month",
    y = "Crude death rate per 100,000",
    title = "Monthly crude death rates by race (Rural Maryland)",
    color = "Race"
  ) +
  theme_minimal()
```

# Modeling


```{r}
month_tab <- precip_month %>%
  left_join(
    temp_month %>%
      dplyr::select(-"ym",-"monthly_deaths" ),
    by = c("year", "month","race"))%>%
  mutate(month_f = factor(month, levels = 1:12, labels = month.abb))

month_tab <- month_tab %>%
  left_join(
    death_month %>%
      dplyr::select(-"monthly_deaths"),
    by = c("ym","race"))

dat_white <- month_tab %>% filter(race == "White")
dat_nonwhite <- month_tab %>% filter(race == "Non-white")
```

## pre pandemic

### White

#### OLS

```{r}
train <- dat_white %>% filter(year <= 1916)
test  <- dat_white %>% filter(ym >= as.Date("1917-01-01") & ym <= as.Date("1918-02-01"))

#### OLS ####
mod_1 <- lm( monthly_deaths ~ month_f , data = train)
mod_2 <- lm( monthly_deaths ~ month_f + log(pop_interp), data = train)
mod_3 <- lm( monthly_deaths ~ month_f + log(pop_interp) + high_temp , data = train)
mod_4 <- lm( monthly_deaths ~ month_f + log(pop_interp) + high_temp + high_precip, data = train)
mod_5 <- lm( monthly_deaths ~ month_f + log(pop_interp) + high_temp * high_precip , data = train)
mod_6 <- lm( monthly_deaths ~  month_f + log(pop_interp) + monthly_mean_temp + high_temp + very_high_temp + 
              monthly_mean_avg_temp + high_avg_temp + very_high_avg_temp + precip + high_precip + very_high_precip, data = train)

mod_7 <- lm( monthly_deaths ~  month_f  + monthly_mean_temp + high_temp + very_high_temp + 
              monthly_mean_avg_temp + high_avg_temp + very_high_avg_temp + precip + high_precip + very_high_precip, data = train)

####### RMSE #######
test <- test %>%
  mutate(pred_mod1 = predict(mod_1, newdata = test, type="response"))

test <- test %>%
  mutate(pred_mod2 = predict(mod_2, newdata = test), type="response")

test <- test %>%
  mutate(pred_mod3 = predict(mod_3, newdata = test, type="response"))

test <- test %>%
  mutate(pred_mod4 = predict(mod_4, newdata = test, type="response"))

test <- test %>%
  mutate(pred_mod5 = predict(mod_5, newdata = test, type="response"))

test <- test %>%
  mutate(pred_mod6 = predict(mod_6, newdata = test, type="response"))

test <- test %>%
  mutate(pred_mod7 = predict(mod_7, newdata = test, type="response"))

rmse <- function(y, yhat) sqrt(mean((y - yhat)^2, na.rm = TRUE))

rmse_mod1 <- rmse(test$monthly_deaths, test$pred_mod1)
rmse_mod2 <- rmse(test$monthly_deaths, test$pred_mod2)
rmse_mod3 <- rmse(test$monthly_deaths, test$pred_mod3)
rmse_mod4 <- rmse(test$monthly_deaths, test$pred_mod4)
rmse_mod5 <- rmse(test$monthly_deaths, test$pred_mod5)
rmse_mod6 <- rmse(test$monthly_deaths, test$pred_mod6)
rmse_mod7 <- rmse(test$monthly_deaths, test$pred_mod7)

rmse_mod1;rmse_mod2;rmse_mod3;rmse_mod4;rmse_mod5;rmse_mod6;rmse_mod7

```

```{r}
###### CV ###
set.seed(8287)
cv_5fold <- trainControl(method = "cv", number = 5)

cv_rmse_mod1 <- train(monthly_deaths ~ month_f , data = train, method = "lm", trControl = cv_5fold)$results$RMSE
cv_rmse_mod2 <- train(monthly_deaths ~ month_f + log(pop_interp) , data = train, method = "lm", trControl = cv_5fold)$results$RMSE
cv_rmse_mod3 <- train(monthly_deaths ~ month_f + log(pop_interp) + high_temp , data = train, method = "lm", trControl = cv_5fold)$results$RMSE
cv_rmse_mod4 <- train(monthly_deaths ~ month_f + log(pop_interp) + high_temp + high_precip , data = train, method = "lm", trControl = cv_5fold)$results$RMSE 
cv_rmse_mod5 <- train(monthly_deaths ~ month_f + log(pop_interp) + high_temp * high_precip , data = train, method = "lm", trControl = cv_5fold)$results$RMSE 
cv_rmse_mod6 <- train(monthly_deaths ~ month_f + log(pop_interp) + monthly_mean_temp + high_temp + very_high_temp + 
              monthly_mean_avg_temp + high_avg_temp + very_high_avg_temp + precip + high_precip + very_high_precip, 
      data = train, method = "lm", trControl = cv_5fold)$results$RMSE

cv_rmse_mod7 <- train(monthly_deaths ~ month_f  + monthly_mean_temp + high_temp + very_high_temp + 
              monthly_mean_avg_temp + high_avg_temp + very_high_avg_temp + precip + high_precip + very_high_precip, 
      data = train, method = "lm", trControl = cv_5fold)$results$RMSE

cv_rmse_mod1;cv_rmse_mod2;cv_rmse_mod3;cv_rmse_mod4;cv_rmse_mod5;cv_rmse_mod6; cv_rmse_mod7

```

#### lasso

```{r}
set.seed(8287)
form <- monthly_deaths ~ month_f  + log(pop_interp) +
  monthly_mean_temp + high_temp + very_high_temp +
  monthly_mean_avg_temp + high_avg_temp + very_high_avg_temp +
  precip + high_precip + very_high_precip

#### lasso ####
x_train <- model.matrix(form, data = train)[, -1]
y_train <- train$monthly_deaths

x_test  <- model.matrix(form, data = test)[, -1]
y_test  <- test$monthly_deaths

cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1, family = "gaussian", nfolds = 5)

lambda_lasso_min <- cv_lasso$lambda.min
lambda_lasso_1se <- cv_lasso$lambda.1se

pred_lasso_min <- predict(cv_lasso, s = lambda_lasso_min, newx = x_test)

cv_rmse_lasso <- sqrt(min(cv_lasso$cvm))
rmse_lasso <- rmse(y_test, pred_lasso_min)
# coef(cv_lasso, s = cv_lasso$lambda.min)

cv_rmse_lasso
rmse_lasso
```

#### Bayesian

##### Poisson 

```{r}
set.seed(8287)
fit_naive <- brm(monthly_deaths ~ month_f + offset(log(pop_interp)),
            data = train, family = poisson())

fit_all <- brm(monthly_deaths ~ month_f  + offset(log(pop_interp)) + monthly_mean_temp + high_temp + very_high_temp + 
              monthly_mean_avg_temp + high_avg_temp + very_high_avg_temp + precip + high_precip + very_high_precip,
            data = train, family = poisson())

fit_selected_poi <- brm(monthly_deaths ~ month_f  + offset(log(pop_interp)) + monthly_mean_avg_temp + very_high_avg_temp,
            data = train, family = poisson())

test <- test %>%
  mutate(pred_mod_b_poi = fitted(fit_naive, newdata = test)[, "Estimate"])

test <- test %>%
  mutate(pred_mod_b_poi_all = fitted(fit_all, newdata = test)[, "Estimate"])

test <- test %>%
  mutate(pred_mod_b_poi_selected = fitted(fit_selected_poi, newdata = test)[, "Estimate"])

rmse_mod_b_poi <- rmse(test$monthly_deaths, test$pred_mod_b_poi)
rmse_mod_b_poi_all <- rmse(test$monthly_deaths, test$pred_mod_b_poi_all)
rmse_mod_b_poi_selected <- rmse(test$monthly_deaths, test$pred_mod_b_poi_selected)

rmse_mod_b_poi; rmse_mod_b_poi_all; rmse_mod_b_poi_selected
```


##### Negative Binomial

```{r}
set.seed(8287)
fit_naive_nb <- brm(monthly_deaths ~ month_f + offset(log(pop_interp)) ,
            data = train, family = negbinomial())

fit_all_nb <- brm(monthly_deaths ~ month_f  + offset(log(pop_interp)) + monthly_mean_temp + high_temp + very_high_temp + 
              monthly_mean_avg_temp + high_avg_temp + very_high_avg_temp + precip + high_precip + very_high_precip,
            data = train, family = negbinomial())

fit_selected_nb <- brm(monthly_deaths ~ month_f  + offset(log(pop_interp)) + monthly_mean_avg_temp + very_high_avg_temp,
            data = train, family = negbinomial())

test <- test %>%
  mutate(pred_mod_b_nb = fitted(fit_naive_nb, newdata = test)[, "Estimate"])

test <- test %>%
  mutate(pred_mod_b_ng_all = fitted(fit_all_nb, newdata = test)[, "Estimate"])

test <- test %>%
  mutate(pred_mod_b_ng_selected = fitted(fit_selected_nb, newdata = test)[, "Estimate"])

rmse_mod_b_nb <- rmse(test$monthly_deaths, test$pred_mod_b_nb)
rmse_mod_b_ng_all <- rmse(test$monthly_deaths, test$pred_mod_b_ng_all)
rmse_mod_b_ng_selected <- rmse(test$monthly_deaths, test$pred_mod_b_ng_selected)

rmse_mod_b_nb;rmse_mod_b_ng_all;rmse_mod_b_ng_selected
```

```{r}
rmse_tab <- data.frame(
  Model    = c("OLS_short", "OLS_full", "Lasso", "b_Poisson_naive", "b_NegBin_naive", "b_Poisson_all", "b_NegBin_all", "b_NegBin_selected", "b_Poisson_selected"),
  CV_RMSE  = round(c(cv_rmse_mod1, cv_rmse_mod6, cv_rmse_lasso, NA, NA, NA, NA, NA, NA), 2),
  Test_RMSE = round(c(rmse_mod1, rmse_mod6, rmse_lasso, rmse_mod_b_poi, rmse_mod_b_nb, rmse_mod_b_poi_all,
                      rmse_mod_b_ng_all,rmse_mod_b_ng_selected, rmse_mod_b_poi_selected), 2)
)

kable(rmse_tab, caption = "5-Fold CV and Test RMSEs")
```

### Non-White 

#### OLS

```{r}
train2 <- dat_nonwhite %>% filter(year <= 1916)
test2  <- dat_nonwhite %>% filter(ym >= as.Date("1917-01-01") & ym <= as.Date("1918-02-01"))

#### OLS ####
mod_1_nw <- lm( monthly_deaths ~ month_f , data = train2)
mod_2_nw <- lm( monthly_deaths ~ month_f + log(pop_interp), data = train2)
mod_3_nw <- lm( monthly_deaths ~ month_f + log(pop_interp) + high_temp , data = train2)
mod_4_nw <- lm( monthly_deaths ~ month_f + log(pop_interp) + high_temp + high_precip, data = train2)
mod_5_nw <- lm( monthly_deaths ~ month_f + log(pop_interp) + high_temp * high_precip , data = train2)
mod_6_nw <- lm( monthly_deaths ~  month_f + log(pop_interp) + monthly_mean_temp + high_temp + very_high_temp +
              monthly_mean_avg_temp + high_avg_temp + very_high_avg_temp + precip + high_precip + very_high_precip, data = train2)

mod_7_nw <- lm( monthly_deaths ~  month_f  + monthly_mean_temp + high_temp + very_high_temp +
              monthly_mean_avg_temp + high_avg_temp + very_high_avg_temp + precip + high_precip + very_high_precip, data = train2)

####### RMSE #######
test2 <- test2 %>%
  mutate(pred_mod1 = predict(mod_1_nw, newdata = test2, type="response"))

test2 <- test2 %>%
  mutate(pred_mod2 = predict(mod_2_nw, newdata = test2), type="response")

test2 <- test2 %>%
  mutate(pred_mod3 = predict(mod_3_nw, newdata = test2, type="response"))

test2 <- test2 %>%
  mutate(pred_mod4 = predict(mod_4_nw, newdata = test2, type="response"))

test2 <- test2 %>%
  mutate(pred_mod5 = predict(mod_5_nw, newdata = test2, type="response"))

test2 <- test2 %>%
  mutate(pred_mod6 = predict(mod_6_nw, newdata = test2, type="response"))

test2 <- test2 %>%
  mutate(pred_mod7 = predict(mod_7_nw, newdata = test2, type="response"))

rmse <- function(y, yhat) sqrt(mean((y - yhat)^2, na.rm = TRUE))

rmse_mod1_nw <- rmse(test2$monthly_deaths, test2$pred_mod1)
rmse_mod2_nw <- rmse(test2$monthly_deaths, test2$pred_mod2)
rmse_mod3_nw <- rmse(test2$monthly_deaths, test2$pred_mod3)
rmse_mod4_nw <- rmse(test2$monthly_deaths, test2$pred_mod4)
rmse_mod5_nw <- rmse(test2$monthly_deaths, test2$pred_mod5)
rmse_mod6_nw <- rmse(test2$monthly_deaths, test2$pred_mod6)
rmse_mod7_nw <- rmse(test2$monthly_deaths, test2$pred_mod7)

rmse_mod1_nw;rmse_mod2_nw;rmse_mod3_nw;rmse_mod4_nw;rmse_mod5_nw;rmse_mod6_nw;rmse_mod7_nw

```

```{r}
###### CV ###
set.seed(8287)
cv_5fold <- trainControl(method = "cv", number = 5)

cv_rmse_mod1_nw <- train(monthly_deaths ~ month_f , data = train2, method = "lm", trControl = cv_5fold)$results$RMSE
cv_rmse_mod2_nw <- train(monthly_deaths ~ month_f + log(pop_interp) , data = train2, method = "lm", trControl = cv_5fold)$results$RMSE
cv_rmse_mod3_nw <- train(monthly_deaths ~ month_f + log(pop_interp) + high_temp , data = train2, method = "lm", trControl = cv_5fold)$results$RMSE
cv_rmse_mod4_nw <- train(monthly_deaths ~ month_f + log(pop_interp) + high_temp + high_precip , data = train2, method = "lm", trControl = cv_5fold)$results$RMSE
cv_rmse_mod5_nw <- train(monthly_deaths ~ month_f + log(pop_interp) + high_temp * high_precip , data = train2, method = "lm", trControl = cv_5fold)$results$RMSE
cv_rmse_mod6_nw <- train(monthly_deaths ~ month_f + log(pop_interp) + monthly_mean_temp + high_temp + very_high_temp +
              monthly_mean_avg_temp + high_avg_temp + very_high_avg_temp + precip + high_precip + very_high_precip,
      data = train2, method = "lm", trControl = cv_5fold)$results$RMSE

cv_rmse_mod7_nw <- train(monthly_deaths ~ month_f  + monthly_mean_temp + high_temp + very_high_temp +
              monthly_mean_avg_temp + high_avg_temp + very_high_avg_temp + precip + high_precip + very_high_precip,
      data = train2, method = "lm", trControl = cv_5fold)$results$RMSE

cv_rmse_mod1_nw;cv_rmse_mod2_nw;cv_rmse_mod3_nw;cv_rmse_mod4_nw;cv_rmse_mod5_nw;cv_rmse_mod6_nw; cv_rmse_mod7_nw

```

#### lasso

```{r}
set.seed(8287)
form <- monthly_deaths ~ month_f  + log(pop_interp) +
  monthly_mean_temp + high_temp + very_high_temp +
  monthly_mean_avg_temp + high_avg_temp + very_high_avg_temp +
  precip + high_precip + very_high_precip

#### lasso ####
x_train2 <- model.matrix(form, data = train2)[, -1]
y_train2 <- train2$monthly_deaths

x_test2  <- model.matrix(form, data = test2)[, -1]
y_test2 <- test2$monthly_deaths

cv_lasso_nw <- cv.glmnet(x_train2, y_train2, alpha = 1, family = "gaussian", nfolds = 5)

lambda_lasso_min2 <- cv_lasso_nw$lambda.min
lambda_lasso_1se2 <- cv_lasso_nw$lambda.1se

pred_lasso_min_nw <- predict(cv_lasso_nw, s = lambda_lasso_min2, newx = x_test2)

cv_rmse_lasso_nw <- sqrt(min(cv_lasso_nw$cvm))
rmse_lasso_nw <- rmse(y_test2, pred_lasso_min_nw)
# coef(cv_lasso_nw, s = cv_lasso_nw$lambda.min)

cv_rmse_lasso_nw
rmse_lasso_nw
```

#### Bayesian

##### Poisson 

```{r}
set.seed(8287)
fit_naive_nw <- brm(monthly_deaths ~ month_f + offset(log(pop_interp)),
            data = train2, family = poisson())

fit_all_nw <- brm(monthly_deaths ~ month_f  + offset(log(pop_interp)) + monthly_mean_temp + high_temp + very_high_temp +
              monthly_mean_avg_temp + high_avg_temp + very_high_avg_temp + precip + high_precip + very_high_precip,
            data = train2, family = poisson())

fit_selected_poi_nw <- brm(monthly_deaths ~ month_f  + offset(log(pop_interp)) + very_high_avg_temp + very_high_precip,
            data = train2, family = poisson())

test2 <- test2 %>%
  mutate(pred_mod_b_poi = fitted(fit_naive_nw, newdata = test2)[, "Estimate"])

test2 <- test2 %>%
  mutate(pred_mod_b_poi_all = fitted(fit_all_nw, newdata = test2)[, "Estimate"])

test2 <- test2 %>%
  mutate(pred_mod_b_poi_selected = fitted(fit_selected_poi_nw, newdata = test2)[, "Estimate"])

rmse_mod_b_poi_nw <- rmse(test2$monthly_deaths, test2$pred_mod_b_poi)
rmse_mod_b_poi_all_nw <- rmse(test2$monthly_deaths, test2$pred_mod_b_poi_all)
rmse_mod_b_poi_selected_nw <- rmse(test2$monthly_deaths, test2$pred_mod_b_poi_selected)

rmse_mod_b_poi_nw; rmse_mod_b_poi_all_nw; rmse_mod_b_poi_selected_nw
```


##### Negative Binomial

```{r}
set.seed(8287)
fit_naive_nb_nw <- brm(monthly_deaths ~ month_f + offset(log(pop_interp)) ,
            data = train2, family = negbinomial())

fit_all_nb_nw <- brm(monthly_deaths ~ month_f  + offset(log(pop_interp)) + monthly_mean_temp + high_temp + very_high_temp +
              monthly_mean_avg_temp + high_avg_temp + very_high_avg_temp + precip + high_precip + very_high_precip,
            data = train2, family = negbinomial())

fit_selected_nb_nw <- brm(monthly_deaths ~ month_f  + offset(log(pop_interp)) + monthly_mean_avg_temp + very_high_avg_temp,
            data = train2, family = negbinomial())

test2 <- test2 %>%
  mutate(pred_mod_b_nb = fitted(fit_naive_nb_nw, newdata = test2)[, "Estimate"])

test2 <- test2 %>%
  mutate(pred_mod_b_ng_all = fitted(fit_all_nb_nw, newdata = test2)[, "Estimate"])

test2 <- test2 %>%
  mutate(pred_mod_b_ng_selected = fitted(fit_selected_nb_nw, newdata = test2)[, "Estimate"])

rmse_mod_b_nb_nw <- rmse(test2$monthly_deaths, test2$pred_mod_b_nb)
rmse_mod_b_ng_all_nw <- rmse(test2$monthly_deaths, test2$pred_mod_b_ng_all)
rmse_mod_b_ng_selected_nw <- rmse(test2$monthly_deaths, test2$pred_mod_b_ng_selected)

rmse_mod_b_nb_nw;rmse_mod_b_ng_all_nw;rmse_mod_b_ng_selected_nw
```

```{r}
rmse_tab_nw <- data.frame(
  Model    = c("OLS_short", "OLS_full", "Lasso", "b_Poisson_naive", "b_NegBin_naive", "b_Poisson_all", "b_NegBin_all", "b_NegBin_selected", "b_Poisson_selected"),
  CV_RMSE  = round(c(cv_rmse_mod1_nw, cv_rmse_mod6_nw, cv_rmse_lasso_nw, NA, NA, NA, NA, NA, NA), 2),
  Test_RMSE = round(c(rmse_mod1_nw, rmse_mod6_nw, rmse_lasso_nw, rmse_mod_b_poi_nw, rmse_mod_b_nb_nw, rmse_mod_b_poi_all_nw,
                      rmse_mod_b_ng_all_nw,rmse_mod_b_ng_selected_nw, rmse_mod_b_poi_selected_nw), 2)
)

kable(rmse_tab_nw, caption = "5-Fold CV and Test RMSEs")
```

## Pandemic

### White 

```{r}
train <- dat_white %>% filter(year <= 1916)
test_pandemic  <- dat_white %>% filter(ym >= as.Date("1918-02-01") & ym <= as.Date("1920-05-01"))

xtest <- model.matrix(form, data = test_pandemic)[, -1]
test_pandemic <- test_pandemic %>%
  mutate(
    pred_ols = predict(mod_1, newdata = test_pandemic, type="response"),
    pred_ols_full = predict(mod_6, newdata = test_pandemic, type="response"), 
    pred_lasso = as.numeric(predict(cv_lasso, s = lambda_lasso_min, newx = xtest)),
    pred_mod_b_poi_naive = fitted(fit_naive, newdata = test_pandemic)[, "Estimate"],
    pred_mod_b_poi_all = fitted(fit_all, newdata = test_pandemic)[, "Estimate"],
    pred_mod_b_nb_naive = fitted(fit_naive_nb, newdata = test_pandemic)[, "Estimate"],
    pred_mod_b_nb_all = fitted(fit_all_nb, newdata = test_pandemic)[, "Estimate"],
    pred_mod_b_nb_selected = fitted(fit_selected_nb, newdata = test_pandemic)[, "Estimate"],
    pred_mod_b_poi_selected = fitted(fit_selected_poi, newdata = test_pandemic)[, "Estimate"]
    )

```


```{r}
excess_df <- test_pandemic %>%
  mutate(
    excess_ols       = monthly_deaths - pred_ols,
    excess_ols_full  = monthly_deaths - pred_ols_full,
    excess_lasso     = monthly_deaths - pred_lasso,
    excess_b_poisson_naive = monthly_deaths - pred_mod_b_poi_naive,
    excess_b_NegBin_naive       = monthly_deaths - pred_mod_b_nb_naive,
    excess_b_poisson       = monthly_deaths - pred_mod_b_poi_all,
    excess_b_NegBin       = monthly_deaths - pred_mod_b_nb_all,
    excess_b_NegBin_selected       = monthly_deaths - pred_mod_b_nb_selected,
    excess_b_poisson_selected       = monthly_deaths - pred_mod_b_poi_selected
  )

total_excess <- summarise(excess_df,
  OLS             = sum(excess_ols,      na.rm = TRUE),
  OLS_full        = sum(excess_ols_full, na.rm = TRUE),
  Lasso           = sum(excess_lasso,    na.rm = TRUE),
  b_poisson_naive       = sum(excess_b_poisson_naive,      na.rm = TRUE),
  b_NegBin_naive       = sum(excess_b_NegBin_naive,      na.rm = TRUE),  
  b_poisson       = sum(excess_b_poisson,      na.rm = TRUE),
  b_NegBin        = sum(excess_b_NegBin,      na.rm = TRUE),
  b_NegBin_selected        = sum(excess_b_NegBin_selected,      na.rm = TRUE),
  b_poisson_selected       = sum(excess_b_poisson_selected,      na.rm = TRUE)
) %>% tidyr::pivot_longer(everything(), names_to="Model", values_to="Estimated_Total_Excess")

kable(total_excess, caption = "Excess Mortality Estimates")%>%
  kable_styling(latex_options = c("hold_position"))
```


```{r}
set.seed(8287)

# predict using Bayesian negative binomial model (naive model - month dummies and population offset only)
epred_nb <- posterior_predict(fit_naive_nb, newdata = test_pandemic)


total_excess_draw_nb <- apply(epred_nb, 1, function(mu) {
  sum(test_pandemic$monthly_deaths - mu)
})

total_excess_mean_nb <- mean(total_excess_draw_nb)
total_excess_ci_nb   <- quantile(total_excess_draw_nb, c(0.025, 0.975))

c(
  mean = total_excess_mean_nb,
  lwr  = total_excess_ci_nb[1],
  upr  = total_excess_ci_nb[2]
)

```

Bayesian negative binomial model (naive: month dummies predictors only)

Estimated total number of pandemic deaths (Whites): 2685.867 (1846.000, 3479.000)

```{r}
race_pandemic_pop <- death_month %>%
  filter(ym >= as.Date("1918-03-01") & ym <= as.Date("1920-05-01")) %>%
  group_by(race) %>%
  summarise(
    pop_sum = sum(pop_interp),
    .groups = "drop"
  )

c(
    rate_mean = total_excess_mean_nb / race_pandemic_pop$pop_sum[2] * 1e5,
    rate_lwr  = total_excess_ci_nb[1]  / race_pandemic_pop$pop_sum[2] * 1e5,
    rate_upr  = total_excess_ci_nb[2]  / race_pandemic_pop$pop_sum[2] * 1e5
  )
```

Estimated rate of pandemic deaths (Whites): 20.669 (14.206, 26.772)

### Non-White 

```{r}
train2 <- dat_nonwhite %>% filter(year <= 1916)
test_pandemic2  <- dat_nonwhite %>% filter(ym >= as.Date("1918-02-01") & ym <= as.Date("1920-05-01"))

xtest2 <- model.matrix(form, data = test_pandemic2)[, -1]
test_pandemic2 <- test_pandemic2 %>%
  mutate(
    pred_ols = predict(mod_1_nw, newdata = test_pandemic2, type="response"),
    pred_ols_full = predict(mod_6_nw, newdata = test_pandemic2, type="response"), ### mod_6 #### 
    pred_lasso = as.numeric(predict(cv_lasso_nw, s = lambda_lasso_min2, newx = xtest2)), ### lasso ###
    # pred_mod_b_poi_naive = predict(fit_naive_nw, newdata = test_pandemic2, re_formula = NA),
    # pred_mod_b_nb_naive = predict(fit_naive_nb_nw, newdata = test_pandemic2, re_formula = NA),
    # pred_mod_b_poi_all = predict(fit_all_nw, newdata = test_pandemic2, re_formula = NA),
    # pred_mod_b_nb_all = predict(fit_all_nb_nw, newdata = test_pandemic2, re_formula = NA),
    # pred_mod_b_nb_selected = predict(fit_selected_nb_nw, newdata = test_pandemic2, re_formula = NA),
    # pred_mod_b_poi_selected = predict(fit_selected_poi_nw, newdata = test_pandemic2, re_formula = NA)
    
    pred_mod_b_poi_naive = fitted(fit_naive_nw, newdata = test_pandemic2)[, "Estimate"],
    pred_mod_b_poi_all = fitted(fit_all_nw, newdata = test_pandemic2)[, "Estimate"],
    pred_mod_b_nb_naive = fitted(fit_naive_nb_nw, newdata = test_pandemic2)[, "Estimate"],
    pred_mod_b_nb_all = fitted(fit_all_nb_nw, newdata = test_pandemic2)[, "Estimate"],
    pred_mod_b_nb_selected = fitted(fit_selected_nb_nw, newdata = test_pandemic2)[, "Estimate"],
    pred_mod_b_poi_selected = fitted(fit_selected_poi_nw, newdata = test_pandemic2)[, "Estimate"]
    )

```


```{r}
excess_df2 <- test_pandemic2 %>%
  mutate(
    excess_ols       = monthly_deaths - pred_ols,
    excess_ols_full  = monthly_deaths - pred_ols_full,
    excess_lasso     = monthly_deaths - pred_lasso,
    excess_b_poisson_naive = monthly_deaths - pred_mod_b_poi_naive,
    excess_b_NegBin_naive       = monthly_deaths - pred_mod_b_nb_naive,
    excess_b_poisson       = monthly_deaths - pred_mod_b_poi_all,
    excess_b_NegBin       = monthly_deaths - pred_mod_b_nb_all,
    excess_b_NegBin_selected       = monthly_deaths - pred_mod_b_nb_selected,
    excess_b_poisson_selected       = monthly_deaths - pred_mod_b_poi_selected
  )

total_excess2 <- summarise(excess_df2,
  OLS             = sum(excess_ols,      na.rm = TRUE),
  OLS_full        = sum(excess_ols_full, na.rm = TRUE),
  Lasso           = sum(excess_lasso,    na.rm = TRUE),
  b_poisson_naive       = sum(excess_b_poisson_naive,      na.rm = TRUE),
  b_NegBin_naive       = sum(excess_b_NegBin_naive,      na.rm = TRUE),  
  b_poisson       = sum(excess_b_poisson,      na.rm = TRUE),
  b_NegBin        = sum(excess_b_NegBin,      na.rm = TRUE),
  b_NegBin_selected        = sum(excess_b_NegBin_selected,      na.rm = TRUE),
  b_poisson_selected       = sum(excess_b_poisson_selected,      na.rm = TRUE)
) %>% tidyr::pivot_longer(everything(), names_to="Model", values_to="Estimated_Total_Excess")

kable(total_excess2, caption = "Excess Mortality Estimates for Non-Whites")%>%
  kable_styling(latex_options = c("hold_position"))
```


```{r}
set.seed(8287)

# predict using Bayesian negative binomial model (all predictors)
epred_nb_nw <- posterior_predict(fit_all_nb_nw, newdata = test_pandemic2)


total_excess_draw_nb_nw <- apply(epred_nb_nw, 1, function(mu) {
  sum(test_pandemic2$monthly_deaths - mu)
})

total_excess_mean_nb_nw <- mean(total_excess_draw_nb_nw)
total_excess_ci_nb_nw   <- quantile(total_excess_draw_nb_nw, c(0.025, 0.975))

c(
  mean_nw = total_excess_mean_nb_nw,
  lwr_nw  = total_excess_ci_nb_nw[1],
  upr_nw  = total_excess_ci_nb_nw[2]
)

```

Bayesian negative binomial model (all predictors)

Estimated total number of pandemic deaths (Non-Whites): 773.290 (94.950, 1303.025)

```{r}
c(
    rate_mean = total_excess_mean_nb_nw / race_pandemic_pop$pop_sum[1] * 1e5,
    rate_lwr  = total_excess_ci_nb_nw[1]  / race_pandemic_pop$pop_sum[1] * 1e5,
    rate_upr  = total_excess_ci_nb_nw[2]  / race_pandemic_pop$pop_sum[1] * 1e5
  )
```

Estimated rate of pandemic deaths (Non-Whites): 24.915 (3.059, 41.983)

```{r}
# export data
# rmse_tab_race <- data.frame(Model = rmse_tab$Model,
#                             CV_RMSE_w = rmse_tab$CV_RMSE,
#                             Test_RMSE_w = rmse_tab$Test_RMSE,
#                             CV_RMSE_nw = rmse_tab_nw$CV_RMSE,
#                             Test_RMSE_nw = rmse_tab_nw$Test_RMS)
# 
# 
# total_excess_race <- data.frame(Model = total_excess$Model,
#                                 Estimated_Total_Excess_w = total_excess$Estimated_Total_Excess,
#                                 Estimated_Total_Excess_nw = total_excess2$Estimated_Total_Excess)

# write.csv(rmse_tab_race, file = 'MD_race_model_eval.csv', row.names = FALSE)
# 
# write.csv(total_excess_race, file = 'MD_race_excess.csv', row.names = FALSE)
```

